\label{sec:conclusion}
We present TrianguLang, a feed-forward framework for language-guided multi-view segmentation and 3D localization that requires no camera calibration or per-scene optimization at inference. Geometry-Aware Semantic Attention (GASA) injects depth-derived geometric bias into cross-view attention, suppressing semantically plausible but geometrically inconsistent correspondences.

Our experiments demonstrate three key findings. First, text-only prompts can surpass click-based methods when combined with geometric cross-view reasoning. Second, the geometric priors learned through GASA transfer strongly across domains: cross-dataset performance more than doubles the best click-based baseline on out-of-domain benchmarks (75.7\% vs.\ 32.2\% on ScanNet++$\to$uCO3D). Third, TrianguLang matches per-scene optimization methods on NVOS and SPIn-NeRF without requiring camera poses, 3D reconstruction, or minutes of per-scene training.

% These results suggest that composing frozen foundation models with lightweight geometric attention is a viable alternative to per-scene optimization. 

Future work includes several directions. \textbf{Scaling training data:} TrianguLang trained on only 230 indoor scenes already surpasses MV-SAM trained on the SA-1B dataset (millions of images) on ScanNet++ evaluation (62.4\% vs.\ 48.9\%), suggesting substantial headroom from scaling to larger and more diverse training corpora; the trend of geometric priors providing disproportionate gains at smaller data scales would likely amplify with additional data. \textbf{Outdoor and open-world deployment:} Extending to outdoor environments \cite{dl3dv} would enable applications such as language-guided geological survey (e.g., segmenting rock formations by type) and text-only spatial reasoning in unstructured scenes without any visual prompts. Additional future directions include ablating alternative VGMs as drop-in depth/pose replacements, temporal consistency for streaming video, combining learned Spatial IDs~\cite{spatialids2026} with metric depth for improved disambiguation, and deploying TrianguLang on a UR10e robot arm for language-guided manipulation tasks.
