Promptable segmentation has developed recently, enabling complex scene understanding through simple visual prompts like text and bounding boxes. SAM \cite{sam} established the foundation by demonstrating promptable segmentation across diverse applications.

Subsequent versions introduced video segmentation \cite{sam2} and text-based querying \cite{sam3}, opening the door for more dynamic localization. However, despite impressive performance, these models lack 3D awareness, making them prone to object flickering and occlusions.


Incorporating geometric awareness into promptable segmentation presents fundamental challenges. Existing geometric representations (e.g. point clouds, voxels, 3DGS \cite{3dgs}, NeRFs \cite{nerf}), enable lifting 2D information into 3D, but require lengthy per-scene optimization and prioritize fidelity over semantic understanding. Additionally, current promptable segmentation models cannot estimate object position in 3D space, while pose estimation methods \cite{conceptpose,foundationpose}, require preexisting masks, bounding boxes, or cropped images as input.
% -challenges: they face long per-scene optimization (i.e. setup time).

%When writing about another work, capture its importance from the perspective of your own study and contributions
Neural rendering has transformed 3D scene reconstruction. NeRF~\cite{nerf} introduced a novel view synthesis technique requiring only sparse input images with known camera poses. 3D Gaussian Splatting~\cite{3dgs} followed, replacing dense volume representations with lighter anisotropic Gaussians and achieving real-time rendering at a fraction of the compute. While impressive, 3DGS retained reliance on calibrated camera poses and did not address downstream tasks such as object detection and semantic scene understanding. Thus, as reliable and contextually-blank backbones for 3D localization and reconstruction, NeRF and 3DGS presented promising foundations for advancing 3D object awareness, segmentation, and scene understanding.
%like NeRF, 3DGS relies on posed image inputs and is not designed for downstream tasks such as object detection or semantic scene awareness).  3DGS does nothing to alleviate NeRF's reliance on multiple image inputs, nor, more importantly, its complete lack of object recognition or intrinsic scene comprehension.


Fusion models have emerged as successful extensions, distilling reconstruction techniques with features from foundation models like CLIP \cite{clip} and SAM \cite{sam}. One approach utilizes natural language grounding to embed high-dimensional language features into 3D space, enabling open-vocabulary querying of reconstructed scenes \cite{lerf, langsplatv2, openscene3d, fmgs}. Identity-based approaches assign semantic attributes to individual scene primitives (e.g., Gaussians), allowing promptable 3D instance segmentation~\cite{gaussiangrouping, SAGA, refersplat}. Spatial reasoning approaches like SpatialRGPT~\cite{spatialrgpt} and 3D-LLM~\cite{3dllm} inject geometric information into large language models for complex spatial queries.
% Another approach relates parts of a scene by assigning identities to the individual building blocks (e.g. the Gaussians), allowing for precise, promptable, 3D instance segmentation and object recognition \cite{gaussiangrouping, SAGA, refersplat}. *Add in VLM approaches such as LEO. Similarly, SpatialRGPT and 3D-LLM inject geometric information into the LLM...

%storing semantic attributes alongside original parameters
%integrating CLIP and DINO into the GS (Langsplat)
%promptable GS using SAM3 (refersplat)

Visual geometric models (VGMs) \cite{da3, pi3, dust3r, VGGT} take a step beyond traditional NeRFs and 3DGS, demonstrating that scene reconstructions can be produced from unposed RGB images in an end-to-end process.
%For 3D localization, monocular depth estimation provides metric depth from a single image \cite{Marigold, da3}.
Nevertheless, these models too lack intrinsic scene awareness, presenting an opportunity for a new framework capable of handling both semantic understanding and 3D localization without per-scene optimization.
% How can we segue into presenting our model?
%VGGT, Pi3, DA3 can all do better reconstructions from single imagw





% You will see papers with preliminaries that provide the mathematical foundation for a topic that is discussed in methodology. That being said, DO NOT put overly complex explanations and/or math in the introduction

% \todo[inline]{Maybe add benefits associated with pointmaps derived due to bijective mapping of pixels to 3d points.}

Bridging these two fields, we propose \textbf{TrianguLang}, a language-guided multi-view segmentation and 3D localization framework. We blend semantic knowledge from SAM3~\cite{sam3} with geometric priors from DA3 through our novel \textbf{Geometry-Aware Semantic Attention (GASA)} decoder, which uses world-space positional encoding and geometric attention biasing to achieve cross-view consistency without explicit correspondence supervision.

% We evaluate TrianguLang on 6 diverse benchmarks:
% \begin{itemize}[nosep,leftmargin=*]
%     \item \textbf{ScanNet++}~\cite{scannet++}: Indoor scenes with 450+ rooms, instance-level annotations
%     \item \textbf{UCo3D}~\cite{UCo3D}: Object-centric 360° captures, 170K videos across 1000 categories
%     \item \textbf{MVImgNet}~\cite{MVImgNet2}: 520k multi-view images of objects
%     \item \textbf{NVOS}~\cite{NVOS}: Multi-view video object segmentation
%     \item \textbf{SpinNeRF}~\cite{spinnerf}: Object-centric captures for part-level evaluation
%     \item \textbf{PartNeXt}~\cite{partnext}: 350K part annotations across 23K models for part-level grounding
% \end{itemize}

\noindent Our contributions include:
\begin{enumerate}[nosep,leftmargin=*]
    \item \textbf{GASA (Geometry-Aware Semantic Attention)}: A novel attention mechanism that combines semantic similarity with geometric constraints from monocular depth, enabling pose-free cross-view consistency without explicit correspondence supervision.

    % \item \textbf{Sheaf-Theoretic Consistency}: A formalization of multi-view semantic agreement as finding a global section of a cellular sheaf, providing a principled regularizer that encourages topological coherence across views.

    \item \textbf{Pose-Free 3D Localization}: Camera-relative object localization via depth unprojection, providing metric 3D coordinates (e.g., ``1.2m ahead, 0.3m left'') without SLAM or camera pose estimation.

    \item \textbf{LLM-Free Spatial Language}: Support for spatial qualifiers and relational queries via direct geometric computation without requiring LLM inference, enabling real-time spatial grounding.
\end{enumerate}


